---
title: "HW 9"
author: "Minci Zhang"
date: '2022-04-16'
output: html_document 


---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning= FALSE)
```

```{r load-pkg, message=FALSE, include=FALSE}
library(brms)
library(dplyr)
library(haven)
library(ggplot2)  # for plots
library(magrittr)  # for `%>%` operator
library(here)
library(readxl)  # for reading excel files
library(modelsummary)  # for summarizing data
library(rstan)
rstan_options(auto_write = TRUE)  # save compiled STAN object
options(mc.cores = 2)  # use two cores
library(posterior)
library(bayesplot)
theme_set(theme_classic() +
    theme(panel.grid.major.y = element_line(color = "grey92")))
```

# Research Question:
1) Would the pattern of gpa predicting changes in popularity over time differ 
for Hispanic vs Asian American students?
2) Would the pattern of gpa predicting changes in likability over time differ 
for Hispanic vs Asian American students?
3) Would the pattern of popularity predicting changes in gpa over time differ 
for Hispanic vs Asian American students?
4) Would the pattern of likability predicting changes in gpa over time differ 
for Hispanic vs Asian American students?




# Variables:
log transformation has been performed on the original gpa1 popularity1 and likability1 values to fix
the skewness of the distributions 

- 'ethnic1final': ethnicity of the students: 1= Hispanic 2=Asian

- 'loggpa1': log transformed gpa at time one 
- 'logpop1': log transformed value for how popular students are at time one 
- 'loglike1': log transformed value for how likable students are at time one 


- 'gpa_d': difference between gpa at time 1 and time 2(one year later)= (gpa2-gpa1)
- 'pop_d': difference between popularity at time 1 and 2= pop2- pop1
- 'like_d': difference between likability at time1 and 2 = like1-like2

## Import Data

```{r data psyc573, message=FALSE, warning=FALSE}
psyc573 <- read_sav("psyc573.sav")

# Recode `ethnic1final` to a factor variable
psyc573$ethnic1final <- psyc573$ethnic1final -1
psyc573$ethnic1final <- factor(psyc573$ethnic1final,
    levels = c(0, 1),
    labels = c("Hispanic", "Asian"))
```

## Variable Summary

```{r summ-var, message=FALSE, warning=FALSE}
# descriptive statistics summary by ethnic groups
datasummary( (gpa_d + pop_d + like_d + logpop1 + loglike1 + loggpa1 )  * 
                (N + Mean + SD + Min + Max + Histogram) ~ 
                factor(ethnic1final, labels = c("Hispanic", "Asian")),
            data = psyc573)

# correlation by ethnic groups
hisp<- psyc573 %>% filter(ethnic1final== 'Hispanic')
asian <- psyc573 %>% filter(ethnic1final == 'Asian')

hisp %>% 
  select(gpa_d, pop_d, like_d, loggpa1, logpop1, loglike1) %>%
  datasummary_correlation(method="pearson")

asian %>% 
  select(gpa_d, pop_d, like_d, loggpa1, logpop1, loglike1) %>%
  datasummary_correlation(method="pearson")
```


# Model for gpa time 1 predicting popularity change score :

## Model

Let $G$ = loggpa1, $P$ = pop_d, $E$ = ethnic1final
$$
 \begin{aligned}
    P_i & \sim N(\mu_i, \sigma)  \\
    \mu_i & = \beta_0 + \beta_1 G_i + \beta_2 E_i + \beta_3 G_i \times E_i \\
  \end{aligned}
$$
## Prior

$$
  \begin{aligned}
    \beta_0 & \sim N(0, 1) \\
    \beta_1 & \sim N(0.1, 0.05) \\
    \beta_2 & \sim N(0.1, 0.05) \\
    \beta_3 & \sim N(0.1, 0.05) \\
    \sigma & \sim t^+_3(0, 2.5)
  \end{aligned}
$$


```{r m1, echo=TRUE, results='hide'}
# gpa1 predicts popularity change score
m1 <- brm(
    pop_d ~ loggpa1 * ethnic1final,
    data = psyc573,
    prior = prior(normal(0.1, 0.05), class = "b") +
        prior(normal(0.1, 0.05), class = "b", coef = "ethnic1finalAsian") +
        prior(normal(0, 1), class = "Intercept") +
        prior(student_t(3, 0, 2.5), class = "sigma"),
    seed = 940,
    iter = 4000
)
 
```

```{r model1 summary, echo=TRUE}
m1
```

```{r posterior check model1, echo=TRUE}
pp_check(m1, type = "dens_overlay_grouped", group = "ethnic1final")
```


# Results:

As shown in the graph below, the chains mixed well.
```{r convergence1, echo=TRUE}
mcmc_rank_hist(m1)
mcmc_trace(as.array(m1))
```


The following table and graph show the posterior distributions of b_hispanic and b_asian
b_loggpa1:ethnic1finalAsian

```{r summ-m1, echo=TRUE}
as_draws(m1) %>%
    mutate_variables(
        b_hisp = b_loggpa1,
        b_asian = b_loggpa1 + `b_loggpa1:ethnic1finalAsian`
    ) %>%
    posterior::subset_draws(
        variable = c("b_hisp", "b_asian")
    ) %>%
    summarize_draws()



plot(
    conditional_effects(m1,
        effects = "loggpa1",
        conditions = data.frame(ethnic1final = c("Hispanic", "Asian"),
                                cond__ = c("Hispanic", "Asian"))
    ),
    points = TRUE
)
```

# Interpretation:

The analysis showed that on average, the patterns for gpa at time 1 predicting changes in levels of
popularity differed for Asian and Hispanic American middle school students. 


Specifically, gpa at time1 is more strongly associated with an increase in popularity from time 1 to 
time2 for Asian American students (posterior mean= 0.11, 90%CI [0.10, 0.30]) than for Hispanic American
students (posterior mean = 0.11, 90% CI [0.03, 0.19]).






# Model for gpa time1 predicting likability change score:

## Models
Let $G$ = loggpa1, $L$ = like_d, $E$ = ethnic1final

$$
 \begin{aligned}
    L_i & \sim N(\mu_i, \sigma)  \\
    \mu_i & = \beta_0 + \beta_1 G_i + \beta_2 E_i + \beta_3 G_i \times E_i \\
  \end{aligned}
$$

## Prior: 
$$
  \begin{aligned}
    \beta_0 & \sim N(0, 1) \\
    \beta_1 & \sim N(0.1, 0.05) \\
    \beta_2 & \sim N(0.1, 0.05) \\
    \beta_3 & \sim N(0.1, 0.05) \\
    \sigma & \sim t^+_3(0, 2.5)
  \end{aligned}
$$

```{r m2, results='hide'}
# gpa1 predicts likability change score
m2 <- brm(
    like_d ~ loggpa1 * ethnic1final,
    data = psyc573,
    prior = prior(normal(0.1, 0.05), class = "b") +
        prior(normal(0.1, 0.05), class = "b", coef = "ethnic1finalAsian") +
        prior(normal(0, 1), class = "Intercept") +
        prior(student_t(3, 0, 2.5), class = "sigma"),
    seed = 941,
    iter = 4000
)
 
```

```{r model2 summary}
summary(m2)
```

```{r posterior check model2}
pp_check(m2, type = "dens_overlay_grouped", group = "ethnic1final")
```

# Results:

As shown in the graph below, the chains mixed well.
```{r convergence2}
mcmc_rank_hist(m2)
mcmc_trace(as.array(m2))
```

The following table and graph show the posterior distributions of b_hispanic and b_asian

```{r summ-m2}
as_draws(m2) %>%
    mutate_variables(
        b_hisp = b_loggpa1,
        b_asian = b_loggpa1 + `b_loggpa1:ethnic1finalAsian`
    ) %>%
    posterior::subset_draws(
        variable = c("b_hisp", "b_asian")
    ) %>%
    summarize_draws()



plot(
    conditional_effects(m2,
        effects = "loggpa1",
        conditions = data.frame(ethnic1final = c("Hispanic", "Asian"),
                                cond__ = c("Hispanic", "Asian"))
    ),
    points = TRUE
)
```


# Interpretation:

The analysis showed that on average, the patterns for gpa at time 1 predicting changes in levels of
likability differed for Asian and Hispanic American middle school students. 


Specifically, gpa at time1 is more strongly associated with an increase in likability from time 1 to 
time2 for Asian American students (posterior mean= 0.21, 90%CI [0.11, 0.32]) than for Hispanic American
students (posterior mean = 0.10, 90% CI [0.02 0.18]).





# Model for pop1 predicting gpa change score :

## Model:
Let $G$ = gpa_d, $P$ = logpop1, $E$ = ethnic1final

$$
 \begin{aligned}
    G_i & \sim N(\mu_i, \sigma)  \\
    \mu_i & = \beta_0 + \beta_1 P_i + \beta_2 E_i + \beta_3 P_i \times E_i \\
  \end{aligned}
$$

## Prior: 
$$
  \begin{aligned}
    \beta_0 & \sim N(1, 0.5) \\
    \beta_1 & \sim N(0.5, 0.5) \\
    \beta_2 & \sim N(0.5, 0.5) \\
    \beta_3 & \sim N(0.5, 0.5) \\
    \sigma & \sim t^+_3(0, 2.5)
  \end{aligned}
$$



```{r m3, results='hide'}
# pop1 predicts gpa change score
m3 <- brm(
    gpa_d ~ logpop1 * ethnic1final,
    data = psyc573,
    prior = prior(normal(0.5, 0.5), class = "b") +
        prior(normal(0.5, 0.5), class = "b", coef = "ethnic1finalAsian") +
        prior(normal(1, 0.5), class = "Intercept") +
        prior(student_t(3, 0, 2.5), class = "sigma"),
    seed = 939,
    iter = 4000
)
 
```


```{r model3 summary}
summary(m3)
```

```{r posterior check model3}
pp_check(m3, type = "dens_overlay_grouped", group = "ethnic1final")
```
# Results:

As shown in the graph below, the chains mixed well.
```{r convergence3}
mcmc_rank_hist(m3)
mcmc_trace(as.array(m3))
```


The following table and graph show the posterior distributions of b_hispanic and b_asian

```{r summ-m3}
as_draws(m3) %>%
    mutate_variables(
        b_hisp = b_logpop1,
        b_asian = b_logpop1 + `b_logpop1:ethnic1finalAsian`
    ) %>%
    posterior::subset_draws(
        variable = c("b_hisp", "b_asian")
    ) %>%
    summarize_draws()



plot(
    conditional_effects(m3,
        effects = "logpop1",
        conditions = data.frame(ethnic1final = c("Hispanic", "Asian"),
                                cond__ = c("Hispanic", "Asian"))
    ),
    points = TRUE
)
```
# Interpretation:

The analysis indicate that the patterns for popularity at time 1 predicting increase in gpa from time1
and time2 did not differ for Hispanic (posterior mean= 0.08, 90 CI [0.00, 0.16]) and Asian (posterior mean= 0.01, 95% CI[-0.00, 0.07]) American students ($\beta_3$ (the beta for pop1 x gpa_d)=-0.017, 95% CI [-0.18 0.05])








# Model for like1 predicting gpa change score :

## Model
Let $G$ = gpa_d, $P$ = loglike1, $E$ = ethnic1final
$$
 \begin{aligned}
    G_i & \sim N(\mu_i, \sigma)  \\
    \mu_i & = \beta_0 + \beta_1 L_i + \beta_2 E_i + \beta_3 L_i \times E_i \\
  \end{aligned}
$$
## Prior

$$
  \begin{aligned}
    \beta_0 & \sim N(1, 0.5) \\
    \beta_1 & \sim N(0.5, 0.5) \\
    \beta_2 & \sim N(0.5, 0.5) \\
    \beta_3 & \sim N(0.5, 0.5) \\
    \sigma & \sim t^+_4(0, 3)
  \end{aligned}
$$




```{r m4, results='hide'}
# like1 predicts gpa change score
m4 <- brm(
    gpa_d ~ loglike1 * ethnic1final,
    data = psyc573,
    prior = prior(normal(0.5, 0.5), class = "b") +
        prior(normal(0.5, 0.5), class = "b", coef = "ethnic1finalAsian") +
        prior(normal(1, 0.5), class = "Intercept") +
        prior(student_t(3, 0, 2.5), class = "sigma"),
    seed = 938,
    iter = 4000
)
 
```

```{r model4 summary}
summary(m4)
```


```{r posterior check mode4}
pp_check(m4, type = "dens_overlay_grouped", group = "ethnic1final")
```



# Results:

As shown in the graph below, the chains mixed well.
```{r convergence4}
mcmc_rank_hist(m4)
mcmc_trace(as.array(m4))
```



The following table and graph show the posterior distributions of b_hispanic and b_asian

```{r summ-m4}
as_draws(m4) %>%
    mutate_variables(
        b_hisp = b_loglike1,
        b_asian = b_loglike1 + `b_loglike1:ethnic1finalAsian`
    ) %>%
    posterior::subset_draws(
        variable = c("b_hisp", "b_asian")
    ) %>%
    summarize_draws()



plot(
    conditional_effects(m4,
        effects = "loglike1",
        conditions = data.frame(ethnic1final = c("Hispanic", "Asian"),
                                cond__ = c("Hispanic", "Asian"))
    ),
    points = TRUE
)
```



# Interpretation:

The analysis indicate that the patterns for likability at time 1 predicting increase in gpa from time1
and time2 did not differ for Hispanic (posterior mean= -0.05, 90 CI [-0.13, 0.03]) and Asian (posterior mean= -0.02, 95% CI[-0.08, 0.03]) American students ($\beta_3$ (the beta for like1 x gpa_d)=0.03, 95% CI [-0.09 0.15])